
```{r}
install.packages('tidyverse')
library(tidyverse)
```
```{r}
install.packages("readr")
library(readr)
```


```{r}
q1_2019 <- read_csv("Divvy_Trips_2019_Q1.csv", show_col_types = FALSE)
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv", show_col_types = FALSE)
```

```{r}
View(q1_2019)
```
To see if the colun match each other, doesn't matter the order need the same

```{r}
colnames(q1_2019)
colnames(q1_2020 )
```
Rename columns to make them consistent with q1_2020 (as this will be the supposed
going-forward table design for Divvy)

```{r}
q1_2019 <- rename(q1_2019
,ride_id = trip_id
,rideable_type = bikeid
,started_at = start_time
,ended_at = end_time
,start_station_name = from_station_name
,start_station_id = from_station_id
,end_station_name = to_station_name
,end_station_id = to_station_id
,member_casual = usertype
)
```

```{r}
str(q1_2019)
str(q1_2020)
```

change the datatype of ride_id & rideable_type
using mutate() Function to add new variables or modify existing variables in a data frame. 

```{r}
q1_2019<- mutate(q1_2019, ride_id = as.character(ride_id),
                 rideable_type = as.character(rideable_type))

str(q1_2019)
```

combine the dataset
```{r}
all_trips <- bind_rows(q1_2019, q1_2020)
View(all_trips)
```

Remove columns
```{r}
all_trips<- all_trips %>%
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "tripduration"))
View(all_trips)
```

Summary of the dataframe
```{r}
colnames(all_trips)
nrow(all_trips)
dim(all_trips)
head(all_trips)
str(all_trips)
summary(all_trips)
```

Check the distinct value in the "member_casual" column
```{r}
unique_member_casual<- unique(all_trips$member_casual)
print(unique_member_casual)
```
There are two names for members and two names for casual riders, need to consolidate that from four to two labels.
```{r}
all_trips <- mutate(all_trips,
                    member_casual = recode(member_casual,
                                           "Subscriber" = "member",
                                           "Customer" = "casual"))
unique_member_casual<- unique(all_trips$member_casual)
print(unique_member_casual) # now only have "member" "casual"values
```

count how many times each category occurs within the member_casual column
```{r}

table(all_trips$member_casual)
```

add some additional columns of data -- such as day, month, year in order to aggregate the data
add the new column called 'date' at the end of data frame, the format is yyyy-mm-dd.
```{r}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
View(all_trips)
```
ectract the month, day, year from the date column 
```{r}
all_trips$month<- format(all_trips$date,"%m")
all_trips$day<- format(all_trips$date,"%d")
all_trips$year<- format(all_trips$date, "%Y")
all_trips$week<- format(all_trips$date, "%A")
```

```{r}
View(all_trips)
```

Calculate the ride duration in seconds
```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
View(all_trips)
```
Inspect the structure of the columns
```{r}
str(all_trips)
```

Convert the 'difftime' num datatype of ride_length column  to numeric so we can run calculations on the data
```{r}
all_trips$ride_length <- as.numeric(all_trips$ride_length)
str(all_trips)
```

create the new data frame to filter out the data set including those that were removed from their docking stations by employees for maintenance or quality checks, not by customers, and also check the data quality to make sure that ride_length should be positive. 

```{r}
all_trip_v2<- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length < 0 ),]
View(all_trip_v2) #788189
```

Conduct the Descriptive Analysis
```{r}
mean(all_trip_v2$ride_length) #average 19.82 minutes
median(all_trip_v2$ride_length) #average 8.98 minutes
max(all_trip_v2$ride_length)  # average 177200 minutes
min(all_trip_v2$ride_length)
```
```{r}
summary(all_trip_v2$ride_length)
```

Compare members and casual users
```{r}
summary_members <- summary(all_trip_v2$ride_length[all_trip_v2$member_casual == "member"])
summary_casuals <- summary(all_trip_v2$ride_length[all_trip_v2$member_casual == "casual"])
summary_members_df <- data.frame(summary_members)
summary_casuals_df <- data.frame(summary_casuals)

# Combine the two summaries into a single data frame for comparison
comparison_df <- data.frame(
  Member = summary_members,
  Casual = summary_casuals
)

# Use View to see the comparison data frame
View(comparison_df)
```
```{r}
aggregate(all_trip_v2$ride_length ~ all_trip_v2$member_casual, FUN = mean)
aggregate(all_trip_v2$ride_length ~ all_trip_v2$member_casual, FUN = median)
aggregate(all_trip_v2$ride_length ~ all_trip_v2$member_casual, FUN = max)
aggregate(all_trip_v2$ride_length ~ all_trip_v2$member_casual, FUN = min)

```

See the average ride time by each day for members vs casual users
```{r}
aggregate(all_trip_v2$ride_length ~ all_trip_v2$member_casual + all_trip_v2$week, FUN = mean)

```

make the days of the week in order
```{r}
all_trip_v2$week<- ordered(all_trip_v2$week, levels= c("Sunday", "Monday",
"Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

Now the day of week are in order. 
```{r}
aggregate(all_trip_v2$ride_length ~ all_trip_v2$member_casual + all_trip_v2$week, FUN = mean)
```

analyse ridership data by type and weekday
```{r}
summary_by_type_weekday <- all_trip_v2 %>%
  group_by(member_casual, week) %>%
  summarise(
    Total_Rides = n(), 
    Average_Duration = mean(ride_length))

View(summary_by_type_weekday)

    
```
```{r}
install.packages("lubridate")

# Load lubridate package
library(lubridate)
install.packages("ggplot2")
library(ggplot2)
```


```{r}

all_trip_v2 %>%
mutate(weekday = wday(started_at, label = TRUE)) %>%
group_by(member_casual, weekday) %>%
summarise(number_of_rides = n()
,average_duration = mean(ride_length)) %>%
arrange(member_casual, weekday) %>%
ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
geom_col(position = "dodge")

#As observed in the graph,  the number of rides by members most gather on weekdays suggests that most are likely commuters. Whereas the number of rides by casual riders increases on Saturdays and Sundays, indicating that these users may include tourists or infrequent users who utilise the service for leisure activities during weekends.
```


```{r}
all_trip_v2 %>%
mutate(weekday = wday(started_at, label = TRUE)) %>%
group_by(member_casual, weekday) %>%
summarise(number_of_rides = n()
,average_duration = mean(ride_length)) %>%
arrange(member_casual, weekday) %>%
ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
geom_col(position = "dodge")

# casual rides on Thursday show a significant deviation from the typical weekly pattern. In contrast, average duration in member remained consistent throughout the week. This suggest Firstly, there is a possibility of promotional strategies on thursday that are specifically targeted at casuals and Secondly,  The consistent average ride time for members indicate that subscription models may encourage shorter, more frequent usage.

```

